\subsection{Lecture 3}
\subsubsection{Concentration Inequalities}

\begin{definition}[Indicator Random Variable]
    $\mathbbm{1}_A$
    is the indicator for event $A$, i.e. a random variable with the following values:

    \[ \mathbbm{1}_A = 
    \begin{cases}
    1 & \text{ if sample point in event $A$} \\
    0 & \text{ otherwise}
    \end{cases} \]
\end{definition}

\begin{theorem} [Markov's Inequality]
    Consider random variable $X \geq 0$ and constant $a > 0$. Then,
    \[ \Pr{X \geq a} \leq \frac{\E{X}}{a} \]
\begin{proof*}
    Let $Y = \mathbbm{1}_{X \geq a}$. Then we know:

    \begin{align*}
        \E{Y} &= 0 \cdot \Pr{X < a} + 1 \cdot \Pr{X \geq a} = \Pr{X \geq a} \\
        Y &\leq \frac{X}{a} \\
        \E{Y} &\leq \E{\frac{X}{a}} = \frac{\E{X}}{a} \\
        \Pr{X \geq a} &\leq \frac{\E{X}}{a}
    \end{align*}
\end{proof*}
\end{theorem}


Markov's inequality tends to be a coarse bound, and $X, a$ have to be non-negative.

\begin{theorem} [Chebyshev's Inequality]
    For random variable $X$ and $\epsilon > 0$:
    \[ \Pr{\abs{X - \E{X}}} \geq \epsilon) \leq \frac{\Var{X}}{\epsilon^2} \]

    Define $Z = \abs{X - \E{X}}^2$, $a = \epsilon^2$, $\epsilon > 0$.

    \begin{proof*}
    Apply Markov's inequality:
    \[ \Pr{Z \geq \epsilon^2} \leq \frac{\E{Z}}{\epsilon^2} \]

    Note that \[\E{Z} = \E{(X - \E{X})^2} = \Var{X} \]

    This means that \[ \Pr{\sqrt{Z} \geq \epsilon} \leq \frac{\Var{X}}{\epsilon^2} \]
    which is exactly the statement of Chebyshev's.
    \end{proof*}
\end{theorem}

Chebyshev's is generally a tighter bound than Markov's.

\begin{theorem}[Weak Law of Large Numbers]
    Assume $X_1, X_2, X_3, \dots$ are independent random variables with the same expectation
    $\mu$ and the same variance $\sigma^2$, and define $Y_n = \frac{(X_1 + X_2 + \dots + X_n)}{n}$
    Then, we have that for any constant $\epsilon > 0$.
    \[ \lim_{n \to \infty} \Pr{\abs{Y_n - \mu} \geq \epsilon} \to 0 \]
\end{theorem}

This can be shown by Chebyshev's inequality, namely note that the expression in the limit
is bounded by $\frac{\Var{Y_n}}{\epsilon^2} = \frac{n\sigma^2}{n^2\epsilon^2} \to 0$.

In words, this means the probability of the sample mean being within $\epsilon$ of the true mean
approaches 1.

\subsubsection{Covariance and Estimation}
\begin{definition} [Covariance]
    \[ \Cov{X, Y} = \E{XY} - \E{X}\E{Y} = \E{(X - \E{X})(Y - \E{Y})} \]
\end{definition}

If $X$ and $Y$ are independent, then $\Cov{X, Y} = 0$. If the latter is true,
then $X$ and $Y$ are uncorrelated. We also define the coefficient of correlation:

\[ \rho_{xy} = \frac{\Cov{X, Y}}{\sigma_X \sigma_Y} \]

This is handy because $\abs{\rho_{XY}} \leq 1$.

Suppose we want to estimate a random variable $Y$ by $\hat{Y}$ given a correlated
random variable $Y$.

We want to minimize $\E{\qty(Y - \hat{Y})^2}$ but have a linear relationship.
This yields LLSE:

\begin{theorem} [Linear Least Squares Estimate]
    The LLSE
    \[ \hat{Y} = \E{Y} + \frac{\Cov{X, Y}}{\Var{X}}\qty(X - \E{X}) \]
    is the best linear estimate of $Y$ given $X$.
\end{theorem}

\section{Basic Probability}

\subsection{Lecture 3, Continued}

\subsubsection{Infinite Collections of Events and Borel-Cantelli}
There are some important consequences of the axioms of probability.

\begin{theorem} [Infinite Sub-Events]
    Consider some set $A$ where $A = \bigcup_{n = 1}^{\infty} A_n$ where:
    \[ A_1 \subseteq A_2 \subseteq \dots \]

    Then, $\Pr{A_n} \rightarrow \Pr{A}$.

    Furthermore, consider some set $B$ where $B = \bigcap_{n = 1}^{\infty} B_n$ where:
    \[ B_1 \supseteq B_2 \supseteq \dots \]
    Then, $\Pr{B_n} \rightarrow \Pr{A}$.
\end{theorem}

\begin{theorem} [Borel-Cantelli Theorem]
    Let $\{A_n\}_{n = 1}^{\infty}$ be a collection of events such that $\sum_{n = 1}^{\infty} \Pr{A_n} < \infty$, then
    $\Pr{A_n \text{ infinitely often}} = 0$.

    $\{A_n \text{ infinitely often}\}$ is the following event:
    \[ \{\omega \mid \exists N(\omega) \text{ such that } \forall n > N(\omega), \omega \not\in A_n \} \]

    In English, the set describes all outcomes where you can assign a number to that outcome such that
    after $A_N$, you have no set membership.

    The theorem claims that you CAN assign such a number (max event) to any outcome with nonzero probability.

    \begin{proof*}
        Define
        \[ B_n = \bigcup_{m \geq n} A_m \]
        Note that $B_1 \supseteq B_2 \supseteq B_3 \dots$
        
        Calling,
        \[ B = \bigcap_{n} B_n = B_n \]
        Note that $\omega \in \{ A_n \text{ io}\}$ if and only if $\omega \in B_n$ for all $n$.
        But this means that $\omega \in B$. This means that $\{ A_n \text{ io}\} = B$.

        So we must calculate $\Pr{B}$. However, note that $\Pr{B_n} \to \Pr{B}$ as $n \to \infty$.
        Thus, we must compute:
        \begin{align*}
            \Pr{B} &= \lim_{n \to \infty} \Pr{B_n} \\
            \Pr{B_n} &\leq \sum_{m = n}^{\infty} \Pr{A_m} \to 0 \\
            \Pr{B} &= 0
        \end{align*}
        The second step is justified by the following result from analysis. For non-negative sequence $a_n$, if $\sum_{i = 1}^{\infty} a_n < \infty$, then
        $\lim_{n \to \infty} \sum_{m = n}^{\infty} a_m \to 0$.

        Our result shows that $\Pr{A_n \text{ io}} = 0$ \qed
    \end{proof*}
\end{theorem}

Consider the following example for coin flips:

\begin{example} [Infinite Coin Flips]
    Consider the experiment of flipping a coin infinitely many times. Let
    \[ A_n = \{ \text{$n$th flip is heads} \} \]
    Then, in this experiment, the event \{$A_n$ infinitely often \} (which we denote as
    $\{A_n \text{ io}\}$)
    \begin{align*}
        \{A_n \text{ io} \} = \{ \omega \mid \text{heads never stop after some $N(\omega)$} \} \\
    \end{align*}
    Here are some sequences that are in that event:
    \begin{align*}
        \omega &= 0, 0, 1, 1, 1, 1, \dots \\
        \omega &= 0, 1, 0, 1, 0, 1, \dots \\
        \omega &= 0, 0, \underbrace{\dots}_{\text{1 million 0's}}, 1, 0, 0, \underbrace{\dots}_{\text{1 million 0's}}, 1, \dots
    \end{align*}
    Now consider the assigning the following probabilities to each heads (instead of the normal, uniform probability space):
    \[ \Pr{A_n} = \frac{1}{n^2} \]
    \(\sum_{n = 1}^{\infty} \frac{1}{n^2}\) converges, so by Borel-Cantelli, $\Pr{A_n \text{ io}} = 0$, i.e.
    the heads ALWAYS stop.

    Now there is one more question. Does $\Pr{A_n \text{ i.o.}} = 0 \implies \{A_n \text{i.o.}\} = 0$?
    The answer is no. In this case, $\Pr{A_n \text{i.o.}} = 0$, but consider the outcome $\omega_n$ where the $n$th
    flip onwards is a heads; these are all in the infinitely often set, so it actually has infinite cardinality!
\end{example}