\subsection{Lecture 8}

\subsubsection{Hitting Times}

\begin{note}
Consider a Markov Chain $X_i$. Suppose we wanted to find
the time it takes to hit any state in some set $A$, starting at some other state $i$.
Formally, we are asking,

\[ \beta(i) := \E{T_A \mid X_0 = i}, i \in \mathcal{X}, A \subseteq \mathcal{X} \]

Since Markov chains are memoryless, we only need to consider the ways to step from
edges out of us, i.e. one step. We call these equations first-step equations (FSEs).
The FSEs for this problem are:
\[
\begin{cases}
    1 + \sum_j P(i, j) \beta(j) & i \notin A \\
    0 & i \in A
\end{cases}
\]

The reasoning for this is as follows. If we are already in set $A$,
it takes us 0 steps to get into it. Otherwise, we take one step to another state $j$,
which has probability $P(i, j)$ of happening, and then the value of $\beta(i)$ will just be $1 + \beta(j)$.
\end{note}

This way of solving for hitting times produces $n^2$ linear equations to solve.

Here are some other problems we can solve with FSEs/Hitting times.

\begin{note}
Suppose you want to find the probability of hitting a state in one set $A$ before another $B$,
\[ \alpha(i) := \Pr{T_A < T_B \mid X_0 = i}, i \in \mathcal{X}, A,B \subseteq \mathcal{X}, A \cap B = \emptyset \]

Then, the FSEs are the following:
\[ 
    \begin{cases}
        1+\sum_j P(i, j) \alpha(j) & i \notin A \cup B \\
        1 & i \in A \\
        0 & i \in B
    \end{cases}
\]

The reasoning is similar: once you hit a bad state, your probability drops to 0 and
if you hit a good state, your probability is 1. Otherwise, it is the sum of the probabilities of
of moving to an adjacent state and that state ending up as something good (the sum is due to total probability rule).
\end{note}

\begin{note}
    Consider a discounted reward for visiting states earlier. Define $h(i)$ as the reward for being
    in state $i$. Then we define
    \[ Z = \sum_{n = 0}^{T_A} \beta^n h(X(n)), A \subseteq \mathcal{X}, 0 < \beta \leq 1 \]
    as the discounted reward (from a gambling run, perhaps). You accrue wealth until you hit $A$,
    but take too long and your reward is reduced. We want to find:
    \[ \E{Z \mid X_0 = i}, i \in X \]
    The FSEs then become:
    \[ \begin{cases}
        h(i) + \beta \sum_j P(i, j) \delta(j) & i \in A \\
        h(i) & i \in A
    \end{cases} \]
\end{note}
