\section{Lecture 11}

\subsection{Describing Noisy Quantum States}
Recall that in a (noisy) classical system, we define a probability distribution over $N$ states
$\{p_i\}_{i = 1}^N$ where each $p_i$ tells you the probability of being in state $i$. Similarly, in a quantum system,
we can define probabilities $p_i$ of being in a state $\ket{\psi_i}$. Suppose we measure in the $\ket{m}$
basis, so the probability of measuring $m$ is
\[ \Pr{m} = \sum_i p_i \Pr{m \mid \psi_i} = \sum_i p_i \qty|\braket*{m}{\psi_i}|^2 \]
But, we can rewrite this quadratic term as a trace:
\[ \abs{\braket{m}{\psi_i}}^2 = \braket*{\psi_i}{m}\braket*{m}{\psi_i} = \Tr\left(\ket{m}\ketbra*{m}{\psi_i} \bra{\psi_i}\right) \]
Using the fact that trace is linear, we can see:
\[ \Pr{m} = \Tr\qty(\qty(\sum_i p_i \braket{\psi_i}{\psi_i}) \ketbra{m}{m}) = \Tr(\rho \ketbra{m}{m})\]
where we define the \textbf{density operator} as:
\[ \rho = \sum_i p_i \ketbra{\psi_i}{\psi_i} \]
Note that $\rho$ consists of $d^2$ matrix elements, where $d$ is the dimension of the Hilbert space.
However, the space of ensembles is infinite-dimensional, because there are an infinite amount of states
and we can define any probability distribution we'd like on them.
The reason for this is becuase different ensembles can have the same $\rho$ oeprator.

\begin{example}
    Take the probability distribution $p(\ket{0}) = 0.5$, $p(\ket{1}) = 0.5$, which gives us density
    \[ \rho = \frac{1}{2} \ketbra{0}{0} + \frac{1}{2} \ketbra{1}{1} = \frac{1}{2} I \]
    But now take the distribution $p(\ket{+}) = 0.5$, $p(\ket{-}) = 0.5$, which gives us the same density
    \[ \rho = \frac{1}{2} \ketbra{+}{+} + \frac{1}{2} \ketbra{-}{-} = \frac{1}{2} I \]
\end{example}

Seeing this, it is natural to say that a density matrix $\rho$ defines a noisy quantum state; if $p_i \neq 0$ for more than one $i$, then
this is
called a \textbf{mixed state}. If you have two systems
with the same density matrix, they are indistinguishable. A state $\ket{\psi} \in \mathcal{H}$ is called a \textbf{pure state}, with density matrix $\rho = \ketbra{\psi}{\psi}$. One may note that this is a \textbf{projector}, i.e. an orthogonal projection matrix which takes a state and projects it onto its component in the $\psi$ direction. Formally:

\begin{definition}[Projector]
    A projector $P$ is a Hermitian matrix with only two eigenvalues, $0$ and $1$, i.e.
    \[ P = P^{\dagger} = P^2 \]
\end{definition}

The nice thing
about this matrix is that we don't need to worry about the global phase indistinguishability, as that will cancel in the
projector expression. Similarly, the projector $\rho$ has some nice properties.

\subsection{Properties of the Density Operator}

We have that
\[ \rho^{\dagger} = \sum_i p_i \qty(\ketbra{\psi_i}{\psi_i})^{\dagger} = \sum_i p_i \ketbra{\psi_i}{\psi_i} = \rho \]
Thus, $\rho$ is Hermitian. Furthermore, it is actually positive semi-definite:
\[ \bra{\psi} \rho \ket{\psi} = \sum_i p_i \qty|\braket{\psi_i}{\psi_i}|^2 \geq 0 \]
Thus, its eigenvalues are all nonnegative. Finally, we have that
\[ \Tr(\rho) = \sum_i p_i = 1 \]

\begin{theorem}
It turns out, any trace-1 PSD matrix $\rho$ is the density matrix of infinitely many
ensembles $\{(p_i, \ket{\psi_i})\}$.

\begin{proof*}
By the spectral theorem, we can write
such a matrix as
\[ \rho = \sum_j \lambda_j \braket{\psi_j}{\psi_j} \]
Since $\rho$ is PSD, then $\lambda_j \geq 0$ and since it has trace 1, $\sum_{i} \lambda_i = 1$.
Thus, this is the density operator for the ensemble $\{(\lambda_j, \ket*{\psi_j})\}$.

By uniqueness of eigendecomposition, this is in fact the unique ensemble in which all the states
are orthonormal.
\end{proof*}
\end{theorem}

Now, let's visit the postulates of Quantum Mechanics in terms of density matrices.
\begin{theorem}
    \begin{enumerate}
    \item \textbf{Superposition}: For pure states, the original superposition principle states that states live in a Hilbert space $\ket{\psi} \in \mathcal{H}$. For density operators,
    we'd say that a state is described as a PSD trace-one matrix $\rho$ acting on $\mathcal{H}$.
    \item \textbf{Measurement}: For pure states, recall that we measure in an orthonormal basis $\{\ket{m}\}$ such that
    $\Pr{m} = \qty|\braket{\psi}{m}|^2$ and $\ket{\psi}$ collapses to $\ket{m}$. But to deal with more general measurements,
    e.g. measuring on a certain qubit, we use a \textbf{projection-valued measurement (PVM)}. We say we have a set of projectors
    $P_m$ such that $\sum P_m = I$. For a measurement on a one-qubit state, we use $P_m = \ketbra{m}{m}$; and for a measurement on the first qubit but not the second,
    we use $P_m = \braket{m}{m} \otimes I$, wherein measurement yields
    \[ \Pr{m} = \mel{\psi}{P_m}{\psi}, \text{ and we update } \ket{\psi} \to \frac{P_m \ket{\psi}}{\sqrt{\mel{\psi}{P_m}{\psi}}} \]
    Now for density operators, measurement turns into
    \[ \Pr{m} = \sum_i p_i \mel{\psi_i}{P_m}{\psi_i} = \Tr\qty(\rho P_m) \]
    The update rule is a bit harder. Measuring $m$ gives information about which $\ket{\psi_i}$ we were in, so we have 
    \[ \rho \to \sum_i \Pr{\ket{\psi_i} \mid m} \frac{P_m \ket{\psi_i} \bra{\psi_i} P_m}{\mel{\psi_i}{P_m}{\psi_i}} \]
    To find the conditional probability that we were in state $\psi_i$ given a measurement of $m$, we compute 
    \[ \Pr{\ket{\psi_i} \mid m} = \frac{p_i \Pr{m \mid \ket{\psi_i}}}{\Pr{m}} = \frac{p_i\mel{\psi_i}{P_m}{\psi_i}}{\Tr(\rho P_m)} \]
    This means the final update is:
    \begin{align*}
        \rho &\to \sum_i \left[\frac{p_i\mel{\psi_i}{P_m}{\psi_i}}{\Tr(\rho P_m)} \cdot \frac{P_m \ket{\psi_i} \bra{\psi_i} P_m}{\bra{\psi_i} P_m \ket{\psi_i}}\right] \\
        &= \frac{P_m \left(\sum_i p_i \ketbra{\psi_i}{\psi_i}\right) P_m}{\Tr(\rho P_m)}\\
        &= \frac{P_m \rho P_m}{\Tr(\rho P_m)}
    \end{align*}
    One can think about this as follows: project the operator $\rho$ into the subspace spanned by $P_m$,
    then ``normalize'' the trace by using the denominator.
    \item \textbf{Unitary Evolution}: For pure states, we evolve a state with a unitary by applying $\ket{\psi} \mapsto U\ket{\psi}$. For mixed states, we have:
    \[ \rho = \sum_i p_i \ketbra{\psi_i}{\psi_i} \mapsto \sum_i p_i U \ketbra{\psi_i}{\psi_i} U^{\dagger} = U \rho U^{\dagger} \]
    \end{enumerate}
\end{theorem}

With that formal stuff out of the way, let's think about a two state system. The spectral decomposition
of such an operator must have only two eigenvectors and with the trace-one condition we have:
\[ \rho = \lambda \ketbra{\psi}{\psi} + (1 - \lambda)(I - \ketbra{\psi}{\psi}) \]
Without loss of generality, we can take $\lambda \geq \frac{1}{2}$ because we can just pick the biggest
we can parametrize $\lambda = \frac{1}{2} \qty(1 + r)$ where $0 \leq r \leq 1$ and we can think of
mixed states geometrically. Recall our parametrization of the Bloch sphere last lecture in terms of $\theta$ and $\phi$;
if $r = 1$ we get a pure state $\ketbra{\psi}{\psi}$, and if $r = 0$, then $\lambda = 1/2$ and the expression at the top is $\rho = \frac{1}{2} I$
regardless of the $\ket{\psi}$ we choose. Other mixed states lie in the inside of the sphere. Any $\rho \propto I$ is called a
``maximally mixed'' state.

\subsection{Reduced Density Matrices}
Suppose you have two coupled systems $\mathcal{H} = \mathcal{H}_A \otimes \mathcal{H}_B$. One can think
of $\mathcal{H}_A$ as a quantum computer or some other quantum system we wish to control; $\mathcal{H}_B$ is any other
noise/system that $\mathcal{H}_A$ can potentially interact with. Given some $\ket{\psi} \in \mathcal{H}_A \otimes \mathcal{H}_B$
we wish to only study its state on $\mathcal{H}_A$. But maybe we can write $\ket{\psi} = \ket{\psi_A} \ket{\psi_B}$—then we'd be very lucky.

What do measurements on $A$ look like? They look like $P_m^{A} \otimes I^B$. The probability of observing $m$ is:
\[ \Pr{m} = \mel{\psi}{P_m^A \otimes I^B}{\psi} = \Tr\qty[\qty(P_m^A \otimes I^B) \ketbra{\psi}{\psi}] \]
Pick some orthonormal bases $\ket{i}_A, \ket{j}_B$ for the two spaces. We then have:
\begin{align*}
    \Pr{m} &= \sum_{i, j} \bra{i}_A \bra{j}_B P_m^A \otimes I^B \ketbra{\psi}{\psi} \ket{i}_A \ket{j}_B \\
    &= \sum_{i} \bra{i}_A P_m^A \qty(\sum_j \bra{j}_B\ket{\psi} \bra{\psi} \ket{j}_B) \ket{i}_A \\
    &= \sum_{i} \bra{i}_A P_m^A \Tr_B(\ketbra{\psi}{\psi}) \ket{i}_A
\end{align*}
where we define
\[ \Tr_B(X) = \sum_j \bra{j}_B X \ket{j}_B. \]
Now, define the \textbf{reduced density matrix} $\rho_A = \Tr_B(\ketbra{\psi}{\psi})$. Then, we have
\[ \Pr{m} = \sum_i \bra{i}_A P_m^A \rho_A \ket{i}_A = \Tr(P_m^A \rho_A) \]
This is exactly our old measurement rule!

\subsubsection{Relationship between Reduced Density Matrices}
Finally, let's think about the relationship between $\rho_A$ and $\rho_B$.
\[ \rho_A = \sum_i \lambda_i \ketbra{\psi_i}{\psi_i} \]
However, $\psi_i$ are orthogonal and form a basis, so we can combine the terms by $\ket{\psi_i}_A$:
\[ \ket{\psi} = \sum_i c_i \ket{\psi_i}_{A} \ket{\phi_i}_B \]
where in general, the $\ket{\phi_i}_B$ are not orthogonal. However, we show that in this construction, they actually are (and we can normalize to make them orthonormal)! %First, we can take out the $c_i$'s to normalize $\braket{\phi_i}{\phi_i} = 1$.
By definition,
\begin{align*}
    \rho_A &= \Tr_B(\ketbra{\psi}{\psi}) \\
    &= \sum_k \bra{k}_B \qty(\sum_{i,j} c_i c_j \ket{\psi_i}_A \ket{\phi_i}_B \bra{\psi_j}_A \bra{\phi_j}_B) \ket{k}_B \\
    &= \sum_{i,j} c_i c_j \ket{\psi_i}_A \bra{\psi_j}_A \qty(\sum_k  \braket{k}{\phi_i}_B \braket{\phi_j}{k}_B)
\end{align*}
Note that the $k$ summation is just taking the inner product $\braket{\phi_j}{\phi_i}_B$.
\begin{align*}
    \rho_A &= \sum_{i, j}  c_i c_j \ketbra{\psi_i}{\psi_j}_A \braket{\phi_j}{\phi_i}_B
\end{align*}
Via the Spectral decomposition, we must have the terms vanish if $i \neq j$, yielding
\[ \braket{\phi_j}{\phi_i}_B = \delta_{ij} \]
Thus the $\phi_i$'s are orthonormal. In general, the \textbf{Schmidt Decomposition} of a state is
\[ \ket{\psi} = \sum_i \sqrt{\lambda_i} \ket{\psi_i}_A \ket{\phi_i}_B \]
where $\psi_i$ and $\phi_i$ are both simultaneously orthonormal. Thus, we must also have:
\[ \rho_B = \sum_i \lambda_i \ket{\phi_i}\bra{\phi_i} \]
meaning $\rho_A$ and $\rho_B$ have the same eigenvalues!
